Fail Faster

### 2025-03-22 paper/paper.tex
blah
3.8

% \begin{figure}
% \begin{lstlisting}
% \end{lstlisting}
% \caption{Int Pair Generators in \bq{} and ScalaCheck}
% \label{fig:overhead-explanation-code}
% \end{figure}

@@

\jwc{Unclear what we should call this section}
\bcp{``Performance opportunities in the RNG?''}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another impediment to finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \bcp{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the controlled experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow randomness library with a faster but semantically equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.

We emphasize that while we believe that the insight that faster sampilng translates to faster bug finding
is a cross-langauge one, the specific technical contents of this section are OCaml-specific.
In the case of most other PBT frameworks \cn, the randomness library used operates on by machine
integers, so this \emph{particular} inefficiency does not exist.

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions---like generating an integer uniformly in a range---may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
times \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently collect in OCaml,
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.
To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.

@@

\begin{figure}[h]
  \includegraphics[scale=0.4]{bind.png}
  \caption{}
  \label{fig:binds}
\end{figure}

>>>>

@@

\jwc{Unclear what we should call this section}
\bcp{``Performance opportunities in the RNG?''}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another impediment to finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \bcp{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the controlled experiment suggested by OCaml's inefficient implementation of
SplitMix. By replacing this relatively slow randomness library with a faster but semantically equivalent
implemetnation, we can precisely quantify the bugfinding speedup that a better randomness library gives across a range of PBT scenarios.
\jwc{Ensure this gets pushed to a top level contrib}

Note that while that the insight that faster sampilng translates to faster bug finding
is a cross-language one, the performance intervention described here is OCaml-specific.
In most other PBT frameworks \cn, the randomness library used operates on machine
integers, so this \emph{particular} inefficiency does not exist.

The precise details of how the SplitMix algorithm works \cn are not important for
present purposes; the key fact is that all of its operations are are
defined in terms of arithmetic and bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory,
which has a significant impact on performance. A single call to
one of the \bq library functions---like generating an arbitrary integer---may call into the \texttt{Splittable\_random}
library times. Each call into \texttt{Splittable\_random}
may sample many times from the core SplitMix sampling routine \texttt{next\_int64}, i.e. using rejection sampling to find a value within a range.
Finally, each call to \texttt{next\_int64} allocates 9 times, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently
collect in OCaml\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}
, we will see in Section~\ref{section:eval} that they can still have
a large performance impact on generators that spend most of their time sampling data.
To avoid this allocation, we
reimplement SplitMix in C and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and
C code\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.}.

@@

\begin{figure}[h]
  \includegraphics[scale=0.4]{bind.png}
  \caption{Relationship between the number of binds---a proxy for reliance on generator library abstractions---and speedup from staging, showing that the two relate linearly.}
  \label{fig:binds}
\end{figure}
